---
title: "mexico_rasters"
author: "Carmen Galaz-García"
date: "5/28/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(raster) ### NOTE: attach this BEFORE tidyverse
library(tidyverse)
library(here)
library(sf)
library(fasterize)
library(patchwork)
```


I generated the raster file *mexico_median.tif* using [Google Earth Engine](https://earthengine.google.com). 
It shows the median of the images in the USGS Landsat 8 Collection 1 (Tier 1 TOA Reflectance) from May 1st 2013 to May 1st 2014. 
To read more on how how I used Google Earth Engine to create this raster file, check out my other blog post on this topic!


# Raster exploration

The Landsat 8 raster I produced using the 


```{r}
landsat_file <- here('mexico_median.tif')

ls_stack <- raster::stack(landsat_file)
plot(ls_stack)
#plotRGB(ls_stack, scale=1.5)
```


```{r}
red <- raster(landsat_file, band = 1) # B4 band = red
green <- raster(landsat_file, band = 2) # B3 band = green
blue <- raster(landsat_file, band = 3) # B2 band = blue

#ls_stack <- raster::stack(landsat_file)
#ls_stack

plot(red, col = hcl.colors(n = 100, palette = 'Reds 2'))
plot(green, col = hcl.colors(n = 100, palette = 'Greens 2'))
plot(blue, col = hcl.colors(n = 100, palette = 'Blues 2'))

```
# Masking

To do the masking and select only the region corresponding to Mexico I used a polygon shapefile made by Hijmans, Robert J. from the Museum of Vertebrate Zoology at UC Berkeley.

```{r}
mex_sf <- read_sf(here('stanford-zc863pb5331-shapefile/zc863pb5331.shp')) %>%
  st_transform(crs(red))   # match CRS

mex_rast <- fasterize::fasterize(mex_sf, red)
plot(mex_rast)
#writeRaster(mex_rast, 'data/county.tif')
```

```{r}
mask(red, mex_rast) %>% plot(col = hcl.colors(n = 100, palette = 'Reds 2'))

```

```{r}
red <- mask(red, mex_rast)*255
blue <- mask(blue, mex_rast)*255
green <- mask(green,mex_rast)*255

#masked_mex <- mask(ls_stack, mex_rast)
#plot(masked_mex)
#plotRGB(ls_stack, scale=1.5)
masked_mex <- raster::stack(red,blue,green)
plot(masked_mex)
plotRGB(masked_mex, scale=160, colNA='skyblue1')
```

# Citation
https://geodata.lib.berkeley.edu/catalog/stanford-zc863pb5331

https://developers.google.com/earth-engine/datasets/catalog/LANDSAT_LC08_C01_T1_TOA
