---
title: "mexico_rasters"
author: "Carmen Galaz-García"
date: "5/28/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(raster) ### NOTE: attach this BEFORE tidyverse
library(tidyverse)
library(here)
library(sf)
library(fasterize)
library(patchwork)
```


I generated the raster file *mexico_median.tif* using [Google Earth Engine](https://earthengine.google.com). 
It is available in this [project's GitHub repo](https://github.com/carmengg/mexico_rasters.git). 
This raster shows the median of the images in the USGS Landsat 8 Collection 1 (Tier 1 TOA Reflectance) from May 1st 2013 to May 1st 2014. 
To read more on how how I used Google Earth Engine to create this raster file, check out my other blog post on this topic!


# Raster exploration

The Landsat 8 raster I produced using Google Earth Engine is a coarse approximation of Mexico's shape. 
It is calculated using all the Landsat 8 scenes that intersect a coarse polygon enclosing Mexico. 
We can see here the red, green and blue layers (RGB).


```{r, include=FALSE}
# --- Reading individual layers
landsat_file <- here('mexico_median_rgb.tif')

red <- raster(landsat_file, band = 1) # B4 band = red
green <- raster(landsat_file, band = 2) # B3 band = green
blue <- raster(landsat_file, band = 3) # B2 band = blue
```


```{r, out.width = "30%", out.height = "30%"}
# -----  EXPLORING RGB LAYERS -----
mex_rgb <- raster::stack(here('mexico_median_rgb.tif')) # read all bands as raster stack
#mex_rgb
# looking at stack information we get that:
#   B4 band = red
#   B3 band = green
#   B2 band = blue

# ---- Define Plot Function ----
color_graph <- function(band,color){
  return (plot(band, 
               col = hcl.colors(n = 100, palette = color),
               axes = FALSE,
               legend = FALSE)
          )
}
# --- Colors
RGB = c('Reds 2', 'Greens 2', 'Blues 2')

# --- plot RGB bands
for (i in 1:3){
  color_graph(mex_rgb[[i]],RGB[i])
}
```

# Masking

To do the masking and select only the region corresponding to Mexico I used a polygon shapefile made by Hijmans, Robert J. from the Museum of Vertebrate Zoology at UC Berkeley. 
To use the polygon shapefile for masking we need to convert it into a raster. 
Then we can use it mask each layer of the raster stack.

```{r}
# --- read shapefile
mex_sf <- read_sf(here('stanford-zc863pb5331-shapefile/zc863pb5331.shp')) %>%
  st_transform(crs(mex_rgb))   # match CRS

# --- transform shapefile into raster using red band as guide
mex_rast <- fasterize::fasterize(mex_sf, mex_rgb[[1]])
#writeRaster(mex_rast, 'data/county.tif') #export mask raster
plot(mex_rast,axes=FALSE, box=FALSE)

```

```{r, out.width = "30%", out.height = "30%"}
# --- mask each layer
for (i in 1:3){
  mex_rgb[[i]] <- mask(mex_rgb[[i]],mex_rast)
}
rm(mex_rast)

# --- plot masked RGB bands
for (i in 1:3){
  color_graph(mex_rgb[[i]],RGB[i])
}

```

# True Color

Now that we have the masked red, green and blue bands, we can use these to make a true color composite image of Mexico.
It is a close replica to what we can see with our eyes. Isn't it pretty? Notice too the storm around Tabasco. This is slightly incovenient because it doesn't allow to have a clear view of the ground.

```{r}

#for (i in 1:3){
#  mex_rgb[[i]] <- mex_rgb[[i]]*255
#}

plotRGB(mex_rgb, scale=0.63, colNA='skyblue1')
```

# NDVI: Normalized Differential Vegetation Index

## Near-Infrared Band
To calculate the NDVI we need the near-infrared band too. 
Since the previous raster only contained RGB bands, here I added a raster I generated with Google Earth Engine that shows the median of the near-infrared band from May 1, 2013 to May 1, 2014.

```{r}
# --- read nir raster
nir <- raster(here('mexico_median_nir.tif')) 

# --- masking (note: had to update mask to match extent of nir raster)
mex_rast2 <- fasterize::fasterize(mex_sf, nir)
nir <- mask(nir, mex_rast2)
rm(mex_rast2)

# --- plot
color_graph(nir,'OrRd')

```

## NDVI calculation

To calculate the NDVI we only need the red and near-infrared bands of the satellite image.
So to avoid indices we can extract the red layer we need. 

```{r}
red <- mex_rgb[[1]]
ndvi <- (nir - red) / (nir + red)
#plot(ndvi, col = hcl.colors(100, 'Grays'), axes=FALSE)
```


```{r}
is_forest <- function(x, thresh = .4) {
  y <- ifelse(x >= thresh, 1, NA)
  return(y)
}

forest <- calc(ndvi, fun = is_forest)
plot(forest, col = 'green4')
```

```{r}
ndvi_df <- raster::rasterToPoints(ndvi) %>%
  as.data.frame()
forest_df <- raster::rasterToPoints(forest) %>%
  as.data.frame()

ggplot(data = ndvi_df, aes(x = x, y = y, fill = layer)) +
  geom_raster() +
  geom_raster(data = forest_df, fill = 'green') +
  coord_sf(expand = 0) +
  scale_fill_gradient(low = 'black', high = 'white') +
  theme_void() +
  theme(panel.background = element_rect(fill = 'slateblue4'))
```


# References
https://geodata.lib.berkeley.edu/catalog/stanford-zc863pb5331

https://developers.google.com/earth-engine/datasets/catalog/LANDSAT_LC08_C01_T1_TOA
